A stomatal model of anatomical tradeoffs between gas exchange and pathogen colonization

### 2020-07-13 ms/ms.tex
r1

\usepackage{longtable}
\usepackage{tikz}

@@

\title[Stomata tradeoff photosynthesis for defense]{A stomatal model of anatomical tradeoffs between photosynthesis and
pathogen defense}
\author[\firstAuthorLast]{\Authors}
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

@@

Stomata evolved to regulate gas exchange in and out of the leaf
(Hetherington and Woodward, 2003; Berry et al., 2010; Chater et al.,
2017), but many plant pathogens take advantage of these chinks in the
leaf cuticular armor to infect prospective hosts (Zeng et al., 2010;
McLachlan et al., 2014; Melotto et al., 2017). The density and size of
stomata set the anatomical maximum rate of stomatal conductance to
CO\(_2\) and water vapor (Brown and Escombe, 1900; Parlange and
Waggoner, 1970; Franks and Farquhar, 2001; Franks and Beerling, 2009b;
Lehmann and Or, 2015; Sack and Buckley, 2016; Harrison et al., 2019),
but the pore area shrinks and expands in response to internal and
external factors to regulate gas exchange dynamically (Buckley, 2019).
Many plant pathogens, including viruses (Murray et al., 2016), bacteria
(Melotto et al., 2006; Underwood et al., 2007), protists (Fawke et al.,
2015), and fungi (Hoch et al., 1987; Zeng et al., 2010) use stomatal
pores to gain entry into the leaf. Since stomatal conductance is a major
limitation on photosynthesis (Farquhar and Sharkey, 1982; Jones, 1985)
while pathogens reduce fitness, this sets up a potential tradeoff
between increased photosynthesis and defense against pathogens. Although
there have been many empirical studies on the effect that pathogens have
on stomata, there is no theoretical framework in which to place these
findings. Lack of a theoreatical framework makes it difficult to answer
general questions about how selection for pathogen defense constrains
maximum rates of gas exchange.

Stomatal anatomy is the key link between gas exchange and pathogen
colonization. The density and size of stomata not only determines the
theoretical maximum stomatal conductance (\gsmax), but is also
proportional to the operational stomatal conductance (\gsop) in many
circumstances (Franks et al., 2009, 2014; Dow et al., 2014a; McElwain et
al., 2016; Murray et al., 2019). Therefore, I use anatomical \gsmax{} as
a proxy for \gsop{} and do not address dynamic changes in stomatal
aperture (see \protect\hyperlink{discussion}{Discussion}). Harrison et
al. (2019) recently reviewed the relationship between stomatal anatomy
and gas exchange in detail.

Many pathogens rely on stomata to gain entry into the leaf, but it is
unclear how anatomical traits that determine \gsmax{} (size and density)
affect the ability of pathogens to colonize leaves. The impact of
pathogens on host fitness is complex, but after a pathogen reaches a
host, the first major step is colonization. Once infected, a pathogen
can reduce fitness. Susceptible hosts can lose much of their biomass or
die, but even resistant hosts must allocate resources to defense or
reduce photosynthesis because of necrosis around sites of infection.
Plants can limit colonization physiologically by closing stomata after
they recognize pathogens, called stomatal defense (Melotto et al.,
2017). Anatomy may be another layer of defense; plants can reduce
pathogen colonization by developing leaves with lower stomatal density
and/or size. Infection increases in leaves with higher stomatal density
(McKown et al., 2014; Tateda et al., 2019; Dutton et al., 2019; Fetter
et al., 2019), which suggests that both anatomical and physiological
responses (stomatal closure) affect host colonization. By the same
logic, if stomatal density were held constant, larger stomata should
also increase colonization because they occupy more area. One key
difference between physiological and anatomical defenses is that
stomatal closure reduces gas exchange transiently, whereas anatomy
constrains gas exchange throught the life of the leaf. Because we do not
understand the relationship between stomatal anatomy and colonization
well, we cannot predict the relationship between \gsmax{} and infection.

If stomatal size and density affect pathogen colonization, selection to
limit colonization may shape stomatal size-density scaling
relationships. Botanists have long recognized that stomatal size and
density are inversely correlated (Weiss, 1865; Tichá, 1982; Hetherington
and Woodward, 2003; Sack et al., 2003; Franks and Beerling, 2009a;
Brodribb et al., 2013; de Boer et al., 2016), but the evolutionary
origin of this relationship is not yet known. Here I argue that
pathogens could shape selection on this relationship. Explanations for
inverse size-density scaling are usually cast in terms of preserving
\gsmax{} and/or total epidermal area allocated to stomata (\fs; de Boer
et al., 2016) because there are many combinations of stomatal size and
density that have same \gsmax{} or same \fs{}:

@@

\(b\) and \(m\) are assumed to be biophysical and morphological
constants, \emph{sensu} Sack and Buckley (2016) (see
\protect\hyperlink{supplementary-material}{Supplementary Material}). If
size and density also affect pathogen colonization, then selection from
foliar pathogens could significantly alter the size-density scaling
relationship. The empirical size-density scaling relationship is linear
on a log-log scale, determined by an intercept \(\alpha\) and slope
\(\beta\):

@@

How would adding pathogens alter these predicted scaling relationships?
For simplicity, imagine two environments, one without foliar pathogens
and one with lots. In the absence of foliar pathogens, we expect
size-density scaling to preserve \gsmax{}, \fs{}, or some least-cost
combination of them. What happens when we introduce pathogens? Assuming
that stomatal size and density increase pathogen colonization, then
selection will favor reduced size and/or density. This would change the
intercept \(\alpha\) but not the slope. The effect of foliar pathogens
on the slope depends on the relationship between size, density, and
probability of colonization. If the probability of colonization is
proportional to the product of \emph{linear} stomatal size
(\(S ^ {0.5}\)) and density (\(\propto D S ^ {0.5}\) as for \gsmax) then
it has the same effect on the slope as \gsmax{} because there are many
combinations of \(D\) and \(S ^ {0.5}\) that have same probability of
colonization. If the probability of colonization is proportional to the
product of \emph{areal} stomatal size (\(S\)) and density
(\(\propto D S\) as for \fs) then it has the same effect on the slope as
\fs{} because there are many combinations of \(D\) and \(S\) that have
same probability of colonization. Alternatively, the probability of
colonization may have a different scaling relationsip (neither 0.5 nor
1) or may be nonlinear on a log-log scale. Unlike \gsmax{} and \fs{}, we
do not have theory to predict a stomatal size-density relationship that
preserves the probability of colonization.

In summary, the physical relationship between stomatal size, density,
and conductance is well established (Harrison et al., 2019). The same
traits likely affect the probability of pathogen colonization, but we do
not have a theoretical model that makes quantitative predictions. The
inverse stomatal size-density relationship has usually been explained in
terms of preserving stomatal conductance and/or stomatal cover, but
selection by pathogens might alter scaling. To address these gaps, the
goals of this study are to 1) introduce a spatially explicit model
pathogen colonization on the leaf surface; 2) use the model to predict
the relationship between \gsmax, \fs{}, and the probability of
colonization; 3) work out what these relationships predict about
stomatal size-density scaling.

@@

In this section, I introduce a spatially explicit model of pathogen
colonization on a leaf surface. I explain the model structure and
assumptions here; the
\protect\hyperlink{materials-and-methods}{Materials and Methods} section
below describes how I analyzed the model to address the goals of the
study. For generality, I refer to a generic ``pathogen'' that lands on
leaf and moves to a stomate. The model is agnostic to the type of
pathogen (virus, bacterium, fungus, etc.) and the specific biological
details of how it moves (biotrophy). For example, motile bacterial cells
can land and move around where fungi may germinate from a cyst and grow
until they form an appresorium for infection. These very different
biotropic movements on the leaf are treated identically here. I used
Sympy version 1.3 (Meurer et al., 2017) for symbolic derivations.

@@

@@

\end{minipage} & \begin{minipage}[b]{0.61\columnwidth}\raggedright

@@

\(D,d\)\strut

@@

\texttt{D,\ d}\strut

@@

\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
stomatal density (\(d = \text{log}~D\))\strut

@@

\(f_s\)\strut

@@

@@

\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright

@@

\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
anatomical maximum stomatal conductance\strut

@@

\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
stomatal radius (\(S = 2 \pi R ^ 2\))\strut

@@

\(S,s\)\strut

@@

\texttt{S,\ s}\strut

@@

@@

\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
stomatal size (\(s = \text{log}~S\))\strut

@@

circumfrence of stomate \(i\)\strut

@@

Stomata develop relatively equal spacing to minimize resistance to
lateral diffusion (Morison et al., 2005), allow space between stomata
(Dow et al., 2014b), and prevent stomatal interference (Lehmann and Or,
2015). Here I assume that stomata are arrayed in an equilateral
triangular grid with a density \(D\) and size (area) \(S\). This
assumption ignores veins, trichomes, and within-leaf variation in
stomatal density. Stomata are therefore arrayed in an evenly spaced grid
(Figure \ref{fig:fig1}a). The interstomatal distance \(U\), measured as
the distance from the center of one stomata to the next, is the maximal
diagonal of the hexagon in \(\mu\textrm{m}\) that forms an equal area
boundary between neighbording stomata. The area of a hexagon is
\(A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2\). By definition the
stomatal density is the inverse of this area, such that
\(D = A^{-1}_\textrm{hexagon} = \frac{2}{\sqrt{3}} U ^ {-2}\).
Therefore, interstomatal distance can be derived from the stomatal
density as:

@@

For example, if the density is
\(D = 10 ^ 2~\textrm{mm}^{-2} = 10 ^ {-4}~\mu\textrm{m}^{-2}\), then
\(U\) is 107.5 \(\mu \textrm{m}\). Parkhurst (1994) described this
result previously. I also make the simplifying assumption that stomata
are perfectly circular with radius \(R\). This may be approximately true
for fully open stomata with kidney-shaped guard cells. Although I assume
stomata are circular here, in calculating \gsmax, I assume typical
allometric relationships between length, width, and pore area (see
\protect\hyperlink{supplementary-material}{Supplementary Material}).

\begin{figure}
  \centering
    \includegraphics[width=6.5in]{../figures/fig1.jpeg}
    \caption{\textbf{A spatially explicit model of stomatal anatomy and pathogen colonization}. \textbf{a}. Stomata are assumed to be in a homogenous equilateral triangular grid, which means that we can extrapolate from \textbf{b.} a focal triangle to the entire leaf. The circles represent idealized stomata; the grey lines between them are for visualization. \textbf{c.} By symmetry, a single focal region within the focal triangle can be modeled and extrapolated to the rest of the area. \textbf{d.} The model assumes that a pathogen, depicted as a grey rod, lands somewhere on the leaf surface and will sucessfully locate a stomate if it moves at the correct angle, depicted by the grey polygons.}
    \label{fig:fig1}
\end{figure}

@@

Now imagine that a pathogen lands at a uniform random position within
the focal region and must arrive at a stomate to colonize. If it lands
on a stomate, then it infects the leaf with probability 1; if it lands
between stomata, then it infects the leaf with probability
\(p_\text{locate}\). This is the probability that it locates a stomate,
which I will derive below. The probabilities of landing on or between a
stomate are \fs{} and \(1 - f_S\), respectively. Hence, the total
probability of colonization is:

@@

I assume that the pathogen cannot sense where stomata are and orients at
random, thereafter traveling in that direction. If it successfully
locates a stomate, it colonizes the leaf, but otherwise does not infect.
If there is a high density of stomata and/or large stomata, the
probability of locating a stomate increases. By assuming that stomata
form an equilateral triangular grid (see above), we can extrapolate what
happens in a focal triangle (Figure \ref{fig:fig1}b) by symmetry.
Further, since an equilateral triangle can be broken up into six
identical units (Figure \ref{fig:fig1}c), we can simply calculate
\(p_\text{locate}\) in this focal area. This implicitly assumes that the
probability of colonzing stomata outside the focal area is 0 because
they are too far away.

Imagine that the pathogens lands in position (\(x_p,y_p\)) within the
triangle. The centroid of the triangle is at position (\(x_c,y_c\)) and
a reference stomate is at position \((0,0)\) (Figure \ref{fig:fig2}a).
Therefore \(x_c = U / 2\) and \(y_c = \sqrt{3} U / 6\). The other
stomata are at positions \((U/2, \sqrt{3} U /2)\) and \((U,0)\) (Figure
\ref{fig:fig2}). \(x_p\) and \(y_p\) are defined as the horizontal and
vertical distances, respectively, from the pathogen to the reference
stomate at position \((0,0)\).

@@

\begin{figure}
\centering
\input{../figures/fig2.tex}
\caption{\textbf{Spatial representation of stomata and pathogen.} \textbf{a.} The pathogen starts at a uniform random position within the focal region denoted ($x_p,y_p$). Within the focal triangle, the reference stomata is at position (0,0) by definition, and other stomatal positions are determined by the interstomatal distance $U$. \textbf{b.} Within the focal region, a pathogen can land within the stomate (white cirlce with grey outline and radius $R$) or in the grey area. The outer borders of this area are shown and depend on $R$ and $U$. For a given position $x$, there is a minimum $y$-value ($y_\text{min}$, dashed line) and maximum $y$-value ($y_\text{max}$, solid line).}
\label{fig:fig2}
\end{figure}

I further assumed that the longer distance a pathogen must travel, the
less likely it would be to locate a stomate. For example, if stomata are
at very low density, then a pathogen may die before it reaches a stomate
because of UV, dessication, or another factor. I included this effect by
assuming the probability of reaching a stomate declines exponentially at
rate \(H\) with the Euclidean distance \(v_i(x_p,y_p)\) between the
pathogen location and the edge of stomata \(i\), which is distance \(R\)
from its center at \(x_i,y_i\):

@@

The probability of locating a stomate as a function of \(x_p\) and
\(y_p\) (\(f_\text{locate}(x_p,y_p)\)) is the sum of the angles divided
by \(2\pi\), discounted by their distance from the stomate:

@@

When \(H = 0\), \(p_\text{locate}\) is the fraction of angles that lead
from \((x_p,y_p)\) to a stomate. When \(H > 0\), \(p_\text{locate}\) is
proportional to this fraction, but less than it depending on stomatal
density, size, and starting location of the pathogen.

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{soul}
\usepackage{tikz}

@@

\title[Stomatal anatomy and pathogen colonization]{A stomatal model of anatomical tradeoffs between gas exchange and
pathogen colonization}
\author[\firstAuthorLast]{\Authors}
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

@@

Stomata evolved to regulate gas exchange in and out of the leaf
(Hetherington and Woodward, 2003; Berry et al., 2010; Chater et al.,
2017), but many foliar pathogens take advantage of these chinks in the
leaf cuticular armor to infect prospective hosts (Zeng et al., 2010;
McLachlan et al., 2014; Melotto et al., 2017). The stomatal and
mesophyll conductance to CO\(_2\) are two major limits to photosynthesis
(Lawson et al., 2018; Flexas et al., 2018) that are partially determined
by stomatal anatomy. Since CO\(_2\) conductance limits photosynthesis
(Farquhar and Sharkey, 1982; Jones, 1985) and pathogen infection can
reduce fitness (Gilbert, 2002), this sets up a potential tradeoff
between increased photosynthesis and defense against pathogens mediated
by stomatal anatomy (McKown et al., 2014; Tateda et al., 2019; Dutton et
al., 2019; Fetter et al., 2019). For example, plants could increase
photosynthetic rate by developing more stomata, but more stomata could
result in more pathogen colonization. The optimal stomatal density,
size, and arrangement on the leaf will depend on the fitness gains from
increased gas exchange and fitness losses imposed by foliar pathogens,
both of which depend on the environment. In the next two paragraphs I
will review the relationship between stomatal anatomy, gas exchange, and
foliar pathogen colonization. Then I will discuss why two anatomical
traits, stomatal size and density, might be crucial components of a
broader tradeoff between photosynthesis and pathogen defense.

The stomatal density and maximum pore area set an anatomical upper limit
on stomatal conductance (Brown and Escombe, 1900; Parlange and Waggoner,
1970; Franks and Farquhar, 2001; Franks and Beerling, 2009b; Lehmann and
Or, 2015; Sack and Buckley, 2016; Harrison et al., 2019), but stomatal
shape, distribution, and patterning also affect gas exchange. Smaller
guard cells and dumbbell-shaped stomata of grasses can respond faster to
environmental changes (Drake et al., 2013), but responsiveness is
further modulated by subsidiary cell anatomy and physiology (Franks and
Farquhar, 2007; Raissig et al., 2017; Gray et al., 2020). Stomatal
clustering reduces gas exchange and photosynthesis because adjacent
stomata interfere with one another (Dow et al., 2014b), diffusion shells
overlap (Lehmann and Or, 2015), and limitations on lateral diffusion of
CO\(_2\) in the mesophyll (Lawson and Blatt, 2014 and references
therein). However, sparse clusters of small stomata could allow a leaf
with low rates of gas exchange to have faster stomatal response compared
to a leaf with large, low-density stomata (Papanatsiou et al., 2017).
Leaves with stomata on both lower and upper surfaces (amphistomatic or
amphistomatous) supply more CO\(_2\) to the mesophyll than hypostomatic
(aka hypostomatous) leaves that only have stomata on the lower surface
(Parkhurst, 1978; Gutschick, 1984; Parkhurst and Mott, 1990; Oguchi et
al., 2018). In addition to anatomy, the pore area shrinks and expands in
response to internal and external factors to regulate gas exchange
dynamically (Buckley, 2019). For example, stomata typically open during
the day and close at night in C\(_3\)/C\(_4\) plants, but the opposite
is true for CAM plants. Shade, high vapor pressure deficits, dry soil
and other factors can cause stomata to (partially) close even in the
middle of the day. Variation in how stomata respond to internal and
external signals may explain as much of the variation in gas exchange
across leaves as anatomy (Lawson and Blatt, 2014).

Many types of foliar pathogens, including viruses (Murray et al., 2016),
bacteria (Melotto et al., 2006; Underwood et al., 2007), protists (Fawke
et al., 2015), and fungi (Hoch et al., 1987; Zeng et al., 2010) use
stomatal pores to gain entry into the leaf. For example, rust fungi
hyphae recognize the angle at which guard cells project from the leaf
surface and use it as a cue for appressorium formation (Allen et al.,
1991). Oomycete pathogens can target open stomata on a leaf (Kiefer et
al., 2002). Plants can limit colonization through innate immunity,
called stomatal defense (recently reviewed in Melotto et al., 2017), by
closing stomata after they recognize microbe-associated molecular
patterns (MAMPs) on pathogen cells. Some bacterial pathogens have
responded by evolving the ability to prevent stomatal closure,
increasing their colonization of the leaf interior (Melotto et al.,
2006). In addition to stomatal closure, anatomical changes in stomatal
density and/or size might provide another layer of defense against
pathogen colonization. For example, infection increases in leaves with
higher stomatal density (McKown et al., 2014; Tateda et al., 2019;
Dutton et al., 2019; Fetter et al., 2019). The positive effect of
stomatal density on infection suggests that infection is limited by the
number or size of locations for colonization, meaning that many
individual pathogens must usually be unable to find stomata or other
suitable locations for colonization. This is actually somewhat
surprising given the ability of some pathogens to search for and sense
stomata (see above).

Stomatal anatomy could be a key link between gas exchange and pathogen
colonization. Although many anatomical factors and stomatal movement
affect gas exchange (see above), here I focus on the density and size of
stomata in a hypostomatous leaf. Stomatal size refers to both the area
of guard cells when fully open, from which one can calculate the pore
area for gas exchange (see \protect\hyperlink{model}{Model}). For
simplicity, I model a hypostomatous leaf, but consider the implications
for amphistomatous leaves in the
\protect\hyperlink{discussion}{Discussion}. Stomatal size and density
not only determine the theoretical maximum stomatal conductance
(\gsmax), but is also proportional to the operational stomatal
conductance (\gsop) in many circumstances (Franks et al., 2009, 2014;
Dow et al., 2014a; McElwain et al., 2016; Murray et al., 2019). \gsop{}
is the actual stomatal conductance of plants in the field and is almost
always below \gsmax{} because stomata are usually not fully open.
Although they are not the same, the strong empirical relationship
between \gsmax{} and \gsop{} means that anatomical \gsmax{} can be used
as a proxy for \gsop{} without explicitly modeling dynamic changes in
stomatal aperture (see \protect\hyperlink{discussion}{Discussion}).
Stomatal size and density have also been measured on many more species
than stomatal responsiveness, which may make it easier to test
predictions.

After a pathogen reaches a host, it must survive on the leaf surface and
colonize the interior (Beattie and Lindow, 1995; Tucker and Talbot,
2001). For analytical tractability, I restrict the focus here to
colonization by a pathogen using a random search on a leaf without
stomatal defense (i.e.~a leaf that cannot recognize pathogens and close
stomata). Obviously, these simplifications ignore a lot of important
plant-pathogen interaction biology. In the
\protect\hyperlink{discussion}{Discussion}, I delve further into these
limitations and suggest future work to overcome these limitations. In
order for pathogen-mediated selection on stomatal anatomy, I assume that
the pathogen reduces host fitness once it colonizes (Gilbert, 2002).
Susceptible hosts can lose much of their biomass or die, but even
resistant hosts must allocate resources to defense or reduce
photosynthesis because of defoliation, biotrophy, or necrosis around
sites of infection (Bastiaans, 1991; Mitchell, 2003).

The purpose of this study is to develop a theoretical framework to test
whether variation in stomatal size and density arises from a tradeoff
between gas exchange and pathogen colonization. Since stomatal size and
density affect both gas exchange and pathogen colonization, selection to
balance these competing demands could shape stomatal size-density
scaling relationships. Botanists have long recognized that stomatal size
and density are inversely correlated (Weiss, 1865; Tichá, 1982;
Hetherington and Woodward, 2003; Sack et al., 2003; Franks and Beerling,
2009a; Brodribb et al., 2013; Boer et al., 2016), but the evolutionary
origin of this relationship is not yet known. Here I argue that
deleterious effects of pathogen infection could shape selection on this
relationship. Explanations for inverse size-density scaling are usually
cast in terms of preserving \gsmax{} and/or stomatal cover (\fs),
defined at the fraction of epidermal area allocated to stomata (Boer et
al., 2016), because there are many combinations of stomatal size and
density that have same \gsmax{} or same \fs{}:

@@

\(D\) and \(S\) are stomatal stomatal density and size, respectively
(see Table 1 for a glossary of mathematical symbols and units). \(b\)
and \(m\) are assumed to be biophysical and morphological constants,
\emph{sensu} Sack and Buckley (2016) (see
\protect\hyperlink{supplementary-material}{Supplementary Material}).
\fs{} is proportional to the more widely used stomatal pore area index
{[}Sack et al. (2003); see
\protect\hyperlink{supplementary-material}{Supplementary Material}{]}.
If size and density also affect pathogen colonization, then selection
from foliar pathogens could significantly alter the size-density scaling
relationship. The empirical size-density scaling relationship is linear
on a log-log scale, determined by an intercept \(\alpha\) and slope
\(\beta\):

@@

How would adding pathogens alter these predicted scaling relationships?
For simplicity, consider two environments, one without foliar pathogens
and one with lots. In the absence of foliar pathogens, we expect
size-density scaling to preserve \gsmax{}, \fs{}, or some least-cost
combination of them. What happens when we introduce pathogens? If
stomatal size and density increase pathogen colonization, then selection
will favor reduced size and/or density. This would change the intercept
\(\alpha\) but not the slope. The effect of foliar pathogens on the
slope depends on the relationship between size, density, and probability
of colonization. If the probability of colonization is proportional to
the product of \emph{linear} stomatal size (\(S ^ {0.5}\)) and density
(\(\propto D S ^ {0.5}\) as for \gsmax) then it has the same effect on
the slope as \gsmax{} because there are many combinations of \(D\) and
\(S ^ {0.5}\) that have same probability of colonization. If the
probability of colonization is proportional to the product of
\emph{areal} stomatal size (\(S\)) and density (\(\propto D S\) as for
\fs) then it has the same effect on the slope as \fs{} because there are
many combinations of \(D\) and \(S\) that have same probability of
colonization. Alternatively, the probability of colonization may have a
different scaling relationship (neither 0.5 nor 1) or may be nonlinear
on a log-log scale. Unlike \gsmax{} and \fs{}, we do not have theory to
predict a stomatal size-density relationship that preserves the
probability of colonization.

In summary, the physical relationship between stomatal size, density,
and conductance is well established (Harrison et al., 2019). Size and
density also likely affect the probability of pathogen colonization, but
we do not have a theoretical model that makes quantitative predictions.
The inverse stomatal size-density relationship has usually been
explained in terms of preserving stomatal conductance and/or stomatal
cover, but selection by pathogens might alter scaling. To address these
gaps, the goals of this study are to 1) introduce a spatially explicit
model pathogen colonization on the leaf surface; 2) use the model to
predict the relationship between \gsmax, \fs, and the probability of
colonization; 3) work out what these relationships predict about
stomatal size-density scaling. I analyzed an idealized, spatially
explicit \protect\hyperlink{model}{Model} of how a pathogen lands on a
leaf and finds a stomate to colonize the leaf using a random search. To
my knowledge, this is the first model that makes quantitative
predictions about the relationship between stomatal anatomy, the
probability of colonization, and their impact on stomatal size-density
scaling.

@@

In this section, I introduce a spatially explicit model of successful
pathogen colonization on a leaf surface. I explain the model structure
and assumptions here; the
\protect\hyperlink{materials-and-methods}{Materials and Methods} section
below describes how I analyzed the model to address the goals of the
study. For generality, I refer to a generic ``pathogen'' that lands on
leaf and moves to a stomate. The model is agnostic to the type of
pathogen (virus, bacterium, fungus, etc.) and the specific biological
details of how it moves. For example, motile bacterial cells can land
and move around (Beattie and Lindow, 1995) whereas fungi may germinate
from a cyst and grow until they form an appresorium for infection
(Tucker and Talbot, 2001). These very different tropic movements on the
leaf are treated identically here. I do not model photosynthesis
explicitly, but assume that stomatal conductance limits carbon fixation,
even though the relationship is nonlinear. I used Sympy version 1.6.1
(Meurer et al., 2017) for symbolic derivations.

@@

\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\raggedright

@@

\end{minipage} & \begin{minipage}[b]{0.59\columnwidth}\raggedright

@@

\(D\)\strut

@@

\texttt{D}\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright

@@

\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
stomatal density\strut

@@

\(d\)\strut
\texttt{d}\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
stomatal density (log-scale, \(d = \text{log}~D\))\strut
\fs\strut

@@

\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright

@@

\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright

@@

\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
theoretical maximum stomatal conductance\strut

@@

\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
stomatal radius (\(S = \pi R ^ 2\))\strut

@@

\(S\)\strut

@@

\texttt{S}\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
stomatal size\strut
\(s\)\strut

@@

\texttt{s}\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright

@@

\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
stomatal size (log-scale, \(s = \text{log}~S\))\strut

@@

circumference of stomate \(i\)\strut

@@

Stomata develop relatively equal spacing to minimize resistance to
lateral diffusion (Morison et al., 2005), allow space between stomata
(Dow et al., 2014b), and prevent stomatal interference (Lehmann and Or,
2015). Here I assume that stomata are arrayed in an equilateral
triangular grid with a density \(D\) and size (area) \(S\) on the
abaxial surface only, since most leaves are hypostomatoues {[}Muir
(2015); but see \protect\hyperlink{discussion}{Discussion}{]}. This
assumption ignores veins, trichomes, and within-leaf variation in
stomatal density. Stomata are therefore arrayed in an evenly spaced grid
(Figure \ref{fig:fig1}a). The interstomatal distance \(U\), measured as
the distance from the center of one stomata to the next, is the maximal
diagonal of the hexagon in \(\mu\textrm{m}\) that forms an equal area
boundary between neighboring stomata. The area of a hexagon is
\(A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2\). By definition the
stomatal density is the inverse of this area, such that
\(D = A^{-1}_\textrm{hexagon} = \frac{2}{\sqrt{3}} U ^ {-2}\).
Therefore, interstomatal distance can be derived from the stomatal
density as:

@@

For example, if the density is
\(D = 10 ^ 2~\textrm{mm}^{-2} = 10 ^ {-4}~\mu\textrm{m}^{-2}\), then
\(U\) is 107.5 \(\mu \textrm{m}\). Parkhurst (1994) described this
result previously. I also make the simplifying assumption that stomata
are perfectly circular with radius \(R\) when fully open. This may be
approximately true for fully open stomata with kidney-shaped guard cells
(Sack and Buckley, 2016 and references therein). Although I assume
stomata are circular here, in calculating \gsmax, I assume typical
allometric relationships between length, width, and pore area {[}Sack
and Buckley (2016); see
\protect\hyperlink{supplementary-material}{Supplementary Material}{]}.

\begin{figure}
  \centering
    \includegraphics[width=6.5in]{../figures/fig1.jpeg}
    \caption{\textbf{A spatially explicit model of stomatal anatomy and pathogen colonization}. \textbf{a}. Stomata are assumed to be in a homogenous equilateral triangular grid, which means that we can extrapolate from \textbf{b.} a focal triangle to the entire leaf. The circles represent idealized stomata; the grey lines between them are for visualization. \textbf{c.} By symmetry, a single focal region within the focal triangle can be modeled and extrapolated to the rest of the triangle. \textbf{d.} The model assumes that a pathogen, depicted as a grey rod, lands somewhere on the leaf surface and will sucessfully locate a stomate if it moves at the correct angle, depicted by the grey polygons.}
    \label{fig:fig1}
\end{figure}

@@

Since stomata are arrayed in a homogeneous grid, we can focus on single
focal triangle (Figure \ref{fig:fig1}b-c). Suppose that an individual
pathogen (e.g.~bacterial cell or fungal spore) lands at a uniform random
position within the focal triangle and must arrive at a stomate to
colonize. If it lands on a stomate, then it infects the leaf with
probability 1; if it lands between stomata, then it infects the leaf
with probability \(p_\text{locate}\). This is the probability that it
locates a stomate, which I will derive below. The probabilities of
landing on or between a stomate are \fs{} and \(1 - f_S\), respectively.
Hence, the total probability of colonization is:

@@

I assume that the pathogen cannot sense where stomata are and orients at
random, thereafter traveling in that direction. If it successfully
locates a stomate, it colonizes the leaf, but otherwise does not infect.
If there is a high density of stomata and/or large stomata, the
probability of locating a stomate increases. By assuming that stomata
form an equilateral triangular grid (see above), we can extrapolate what
happens in the focal triangle (Figure \ref{fig:fig1}b) by symmetry.
Further, since an equilateral triangle can be broken up into six
identical units (Figure \ref{fig:fig1}c), we can simply calculate
\(p_\text{locate}\) in this focal area. This implicitly assumes that the
probability of colonizing stomata outside the focal area is 0 because
they are too far away. This assumption may be unrealistic for larger
pathogens, such as fungi, whose hyphae can travel longer distances on
the leaf surface (Brand and Gow, 2012). In
\protect\hyperlink{appendix-1-spatially-implicit-model}{Appendix 1:
Spatially implicit model} I derive a simpler, but spatially
\emph{implicit} model that relaxes the assumption the pathogens must
colonize a stomate within their focal triangle.

Consider a pathogen that lands in position (\(x_p,y_p\)) within the
triangle. The centroid of the triangle is at position (\(x_c,y_c\)) and
a reference stomate is at position \((0,0)\) (Figure \ref{fig:fig2}a).
Therefore \(x_c = U / 2\) and \(y_c = \sqrt{3} U / 6\). The other
stomata are at positions \((U/2, \sqrt{3} U /2)\) and \((U,0)\) (Figure
\ref{fig:fig2}). \(x_p\) and \(y_p\) are defined as the horizontal and
vertical distances, respectively, from the pathogen to the reference
stomate at position \((0,0)\).

@@

\begin{figure}
\centering
\input{../figures/fig2.tex}
\caption{\textbf{Spatial representation of stomata and pathogen.} \textbf{a.} The pathogen starts at a uniform random position within the focal region denoted ($x_p,y_p$). Within the focal triangle, the reference stomate is at position (0,0) by definition, and other stomatal positions are determined by the interstomatal distance $U$. \textbf{b.} Within the focal region, a pathogen can land within the stomate (white circle with grey outline and radius $R$) or in the grey area. The outer borders of this area are shown and depend on $R$ and $U$. For a given position $x$, there is a minimum $y$-value ($y_\text{min}$, dashed line) and maximum $y$-value ($y_\text{max}$, solid line).}
\label{fig:fig2}
\end{figure}

I further assumed that the longer distance a pathogen must travel, the
less likely it would be to locate a stomate. For example, if stomata are
at very low density, then a pathogen may die before it reaches a stomate
because of UV, desiccation, or another factor. I included this effect by
assuming the probability of reaching a stomate declines exponentially at
rate \(H\) with the Euclidean distance \(v_i(x_p,y_p)\) between the
pathogen location and the edge of stomata \(i\), which is distance \(R\)
from its center at \(x_i,y_i\):

@@

The probability of locating a stomate as a function of pathogen position
(\(x_p\) and \(y_p\)) is the sum of the angles divided by \(2\pi\),
discounted by their distance from the stomate:

@@

When there is no pathogen death (\(H = 0\)), \(p_\text{locate}\) is the
fraction of angles that lead from \((x_p,y_p)\) to a stomate. When
\(H > 0\), \(p_\text{locate}\) is proportional to this fraction, but
less than it depending on stomatal density, size, and starting location
of the pathogen.

To obtain the average \(p_\text{locate}\), we must integrate
\(f_\text{locate}(x_p,y_p)\) over all possible starting positions
\((x_p,y_p)\) within the focal area. The focal area is a 30-60-90
triangle with vertices at the center of the reference stomate \((0,0)\),
the midpoint of baseline \((U/2,0)\), and the centroid of the focal
triangle \((U/2, \sqrt{3} / 6 U)\) (Figure \ref{fig:fig1}c).
Colonization occurs with probability 1 if the pathogen lands in the
reference stomate, so we need to integrate the probability of
colonization if it lands elsewhere. This region extends from the edge of
the stomate, at \(\sqrt{3} / 2 R\) to \(U/2\) (Figure \ref{fig:fig2}b).
At any \(x\), we integrate from the bottom of the focal area
(\(y_\text{min}\)) to the top (\(y_\text{max}\)):
